---
title: "Getting Started with ml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
has_ranger <- requireNamespace("ranger", quietly = TRUE)
```

```{r setup}
library(ml)
```

## Why another ML package?

Most machine learning code leaks information from test data into training
without anyone noticing. Preprocessing on the full dataset, repeatedly
evaluating on a single held-out set, using the same data to select
hyperparameters and report final performance -- each of these inflates
reported accuracy and produces models that disappoint in production.

Chapter 7 of Hastie, Tibshirani, and Friedman (2009) prescribes a strict
separation: a **validation** set for iterating on model choices, and a
**test** set touched exactly once for the final generalization estimate.
The `ml` package enforces that separation at the API level.

The workflow has four verbs:

1. **Split** the data into train, validation, and test.
2. **Fit** a model on training data.
3. **Evaluate** on validation data -- repeat as often as you like.
4. **Assess** on test data -- once, when you are done.

Calling `ml_evaluate()` ten times is fine. Calling `ml_assess()` ten times
defeats the purpose. The API makes the distinction explicit.

## Profile

Before modeling, understand what you have:

```{r profile}
prof <- ml_profile(iris, "Species")
prof
```

## Split

Three-way split: 60% train, 20% validation, 20% test. Classification
targets are stratified automatically.

```{r split}
s <- ml_split(iris, "Species", seed = 42)
s
```

The `$dev` property merges train and validation for final retraining before
deployment. Use it only after you have settled on an algorithm and
hyperparameters.

## Screen

Which algorithms work best on this data? `ml_screen()` fits every available
algorithm and ranks them by validation performance.

```{r screen, eval = has_ranger}
lb <- ml_screen(s, "Species", seed = 42)
lb
```

The leaderboard shows the best algorithm at the top. But the winning
algorithm benefits from selection bias (it looked best on *this* particular
validation set). Use `ml_validate()` on test data for trustworthy numbers.

## Fit and evaluate

Train a model and check it on the validation set. You can call
`ml_evaluate()` as many times as you want -- that is the whole point of
having a separate validation partition.

```{r fit-evaluate}
model <- ml_fit(s$train, "Species", algorithm = "logistic", seed = 42)
model

metrics <- ml_evaluate(model, s$valid)
metrics
```

## Visual diagnostics

`ml_plot()` produces base R graphics with no extra dependencies.

```{r plot-importance, fig.height = 3.5}
ml_plot(model, kind = "importance")
```

```{r plot-confusion, fig.height = 4}
ml_plot(model, data = s$valid, kind = "confusion")
```

## Explain

Feature importances normalized to sum to 1.0. For logistic regression these
are absolute coefficient magnitudes; for tree-based models, impurity-based
importances.

```{r explain}
ml_explain(model)
```

## Validate

Gate your model before the final assessment. Define thresholds that must be
met. If any rule fails, `$passed` is `FALSE` and the report shows which
rules were violated.

```{r validate}
gate <- ml_validate(model,
                    test  = s$test,
                    rules = list(accuracy = ">0.80"))
gate
```

## Assess (once)

The test set is your final exam. Call `ml_assess()` exactly once, when you
are done experimenting. The result is your honest generalization estimate.

```{r assess}
verdict <- ml_assess(model, test = s$test)
verdict
```

## Drift detection

When new data arrives, check whether the input distribution has shifted.
No labels required -- this uses per-feature statistical tests.

```{r drift}
# Simulate drift: shift one feature by 2 units
drifted <- s$test
drifted$Sepal.Length <- drifted$Sepal.Length + 2

result <- ml_drift(reference = s$train, new = drifted, target = "Species")
result
```

## Save and load

Models serialize to a compact `.mlr` format:

```{r io}
path <- file.path(tempdir(), "iris_model.mlr")
ml_save(model, path)
loaded <- ml_load(path)
identical(predict(model, s$valid), predict(loaded, s$valid))
```

## Two API styles

Every function is available both as `ml_verb()` (idiomatic R) and as
`ml$verb()` (module style, mirrors the Python package):

```{r module-style}
m2 <- ml$fit(s$train, "Species", algorithm = "logistic", seed = 42)
identical(predict(model, s$valid), predict(m2, s$valid))
```

## Regression

The same workflow applies to continuous targets. The task is detected
automatically from the target column type.

```{r regression, eval = has_ranger}
s2    <- ml_split(mtcars, "mpg", seed = 42)
m_reg <- ml_fit(s2$train, "mpg", seed = 42)
ml_evaluate(m_reg, s2$valid)
```

```{r regression-plot, eval = has_ranger, fig.height = 4}
ml_plot(m_reg, data = s2$valid, kind = "residual")
```

## Available algorithms

```{r algorithms}
ml_algorithms()
```

## References

Hastie, T., Tibshirani, R. and Friedman, J. (2009). *The Elements of
Statistical Learning*, 2nd edition, Springer. Chapter 7: Model Assessment
and Selection.
